---
title: "STAT 405 Project Proposal: Bayesian Hierarchical Modeling of Cancer Mutation Rates"
author: "Torrin Pataki, Annmarie Thomson"
date: "2026-03-16"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(readr)
library(cbioportalR)
library(dplyr)
library(tidyr)
```

## Team

Torrin Pataki and Annmarie Thomson

**Repository**: [github.com/annmariet/stat405_project](https://github.com/annmariet/stat405_project)

## Research Question

Somatic mutation rates for cancer-driving genes vary dramatically across cancer types. For example, the gene *TP53* is mutated in \~95% of ovarian cancers but only \~2% of kidney cancers.

Our research question is: How much do cancer risks and mutation patterns vary across cancer types, and what common structures can be learned by borrowing information across related cancers?

## Main Dataset: TCGA PanCancer Atlas via cBioPortal

The Cancer Genome Atlas (TCGA) PanCancer Atlas provides uniformly processed somatic mutation data across 33 cancer types and \~11,000 tumors. We access it through the public cBioPortal API using the `cbioportalR` R package.

**URL**: <https://www.cbioportal.org/>

We will extract mutation counts for 10 well-studied oncogenes and tumor suppressors (TP53, KRAS, PIK3CA, BRAF, PTEN, APC, ARID1A, EGFR, IDH1, CDKN2A) across \~20 cancer types with $N \geq 150$ sequenced samples. Each observation is a (gene, cancer type) pair with counts $(y_{gc}, n_{gc})$, where $y_{gc}$ is the number of samples harboring a mutation in gene $g$ in cancer type $c$, and $n_{gc}$ is the total number of sequenced samples.

```{r}
set_cbioportal_db("public")

# pull mutation data for 3 cancer types as prototype
study_ids <- c("brca_tcga_pan_can_atlas_2018",
               "luad_tcga_pan_csan_atlas_2018",
               "skcm_tcga_pan_can_atlas_2018")

genes <- c("TP53", "KRAS", "PIK3CA", "BRAF", "PTEN")

pull_counts <- function(sid, genes) {
  # get sample list for this study
  samples <- available_samples(study_id = sid)
  n_total <- nrow(samples)

  # query mutations for target genes only (avoids downloading entire study)
  muts <- get_mutations_by_sample(
    sample_id = samples$sampleId,
    study_id = sid,
    genes = genes
  )

  counts <- muts %>%
    distinct(hugoGeneSymbol, sampleId) %>%
    count(hugoGeneSymbol, name = "n_mutated")

  tibble(gene = genes) %>%
    left_join(counts, by = c("gene" = "hugoGeneSymbol")) %>%
    mutate(n_mutated = replace_na(n_mutated, 0),
           n_total = n_total,
           study = sid,
           cancer = gsub("_tcga_pan_can_atlas_2018", "", sid),
           rate = round(n_mutated / n_total, 3))
}

prototype <- bind_rows(lapply(study_ids, pull_counts, genes = genes))
```

```{r show-data}
head(prototype, 15)
str(prototype)
```

The full dataset will expand this to 10 genes $\times$ \~20 cancer types $\approx$ 200 observations.

## Backup Dataset: The Catalogue of Somatic Mutations in Cancer (COSMIC) Datasets

If the TCGA/cBioPortal approach encounters issues, we will pivot to the COSMIC (Catalogue Of Somatic Mutations In Cancer) database, a large curated resource of somatic mutation information across human cancers.

**URL**: <https://cancer.sanger.ac.uk/cosmic/download/cosmic>

```{r}
mutant_census <- read_tsv(
  "Cosmic_MutantCensus_v103_GRCh37.tsv",
  n_max = 10
)

head(mutant_census)
```

## Methodology plan

**Model 1 (naive)**: Independent Beta-Binomial models per gene per cancer type. $$y_{gc} \mid \theta_{gc} \sim \text{Binomial}(n_{gc},\, \theta_{gc}), \quad \theta_{gc} \sim \text{Beta}(1, 1)$$

**Model 2 (hierarchical)**: Hierarchical Beta-Binomial with gene-level hyperparameters. $$y_{gc} \mid \theta_{gc} \sim \text{Binomial}(n_{gc},\, \theta_{gc}), \quad \theta_{gc} \sim \text{Beta}(\alpha_g,\, \beta_g)$$ $$\alpha_g,\, \beta_g \sim \text{hyperprior}$$

This enables partial pooling of mutation rates across cancer types within each gene.

**Additional analyses**: prior sensitivity analysis (varying hyperprior choices), posterior predictive checks, and comparison of shrinkage behavior between the naive and hierarchical models.

### Team Plan

To ensure that the work is divided equally, we have created a team schedule and will meet at least once a week to make sure that we are consistently working towards deadlines.

| Week | Tasks |
|----|----|
| Mar 8 - Mar 14 | Determine data sets, identify research questions, and turn in proposal |
| Mar 15 - Mar 21 | Work on literature review, develop research questions, identify gaps |
| Mar 22 - Mar 28 | Determine naiive method and work on analysis, work on interpretation |
| Mar 29 - Apr 4 | Determine more complex method(s) and work on analysis and interpretation |
| Apr 5 - Apr 11 | Add discussion, conclusions, implications |
| Apr 12 - Apr 17 | Finish report and submit |
